Variana Reloaded

### 2024-02-06 cv/cv.tex
Work on CV

\noindent Hands-on technical team leader, engineer at heart, PhD, with a specialist expertise in computer vision and artificial intelligence. Has operated in public research and private R\&D in various fields (healthcare, electro-domestic, entertainment). Strong experience in the development of both open source and commercial software in Python, C, C++. {\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, computer-aided diagnosis, smart appliances, digital avatars.

@@

\ \\
\noindent \textbf{SCIENTIFIC REPORTING:} Main author of 25 peer-reviewed scientific articles published in international journals, proceedings and books, co-author of more than 100 scholarly publications and 5 published international patents.

@@

\bighead{HEAD OF ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
\nopagebreak
\begin{itemize}
\item Leading development of generation pipeline of high-fidelity avatars from photos. {\em Keywords:} facial reconstruction, deep face, animation retargeting, image feature extraction, image restoration.
\end{itemize}
\ \\

@@

\bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
\nopagebreak
\begin{itemize}
\item Led algorithmic development of a brain morphometry tool (released as part of the {\em AI-Rad Companion} Siemens solutions) to help radiological reading for patients with suspected neurodegeneration. {\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, machine learning for disease classification. 
\item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
\end{itemize}

@@

\begin{tabular}{p{0.2\linewidth}p{0.8\linewidth}}
  \textbf{Languages:} & Native French, fluent English, basic Portuguese \& Italian.\\
  %\hline
  \ \\
  \textbf{IT skills:} &
  \tabitem Programming languages: Python, C, C++, Matlab, R.\\
  & \tabitem Long-term experience with scientific Python packages (numpy, scipy, pylab).\\
  & \tabitem CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
  & \tabitem  C/Python integration via Cython.\\
  & \tabitem Version control software: git, svn, perforce.\\
  & \tabitem Agile workflow (mainly Scrum) using JIRA \& Confluence.\\
  & \tabitem Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs.\\
  %\hline
  \ \\
  \textbf{Publications:} &
  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}. \\
  & \tabitem 116 scholarly publications.\\
  & \tabitem Google scholar statistics: 7603 citations, h-index: 35, i10-index: 74 (May 2023).\\
  & \tabitem 5 published international patents. 2 ongoing applications.\\
  %\hline
  \ \\
  \textbf{Academic work:} & 
  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (France: Ecole Centrale Paris, University of Nice, INSERM/CNRS continuing education; Switzerland: EPFL Lausanne, University of Lausanne).\\
  & \tabitem Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\\
  & \tabitem Supervision of two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
  %\hline
  \ \\
  \textbf{References:} & Available on request.
\end{tabular}

>>>>

\noindent Hands-on technical team leader, applied researcher and engineer at heart, PhD, with a specialist expertise in computer vision and artificial intelligence. Has operated in public research and private R\&D in various fields (healthcare, electro-domestic, entertainment). Strong experience in the development of both open source and commercial software. {\em Keywords:} applied mathematics, statistics, machine learning, computer vision, computer graphics, computer-aided diagnosis, smart appliances, digital avatars, Python, C.

@@

\ \\
\noindent \textbf{SCIENTIFIC REPORTING:} Author of more than 100 peer-reviewed scientific articles published in international journals, proceedings and books, and 6 published international patents.

@@

\bighead{HEAD OF ARTIFICIAL INTELLIGENCE}{Didimo, Porto, Portugal}{2019--present}
\nopagebreak
\begin{itemize}
\item Leading development of generation pipeline of high-fidelity avatars from photos. {\em Keywords:} facial reconstruction, deep face, animation retargeting, image processing. Accepted talk at SIGGRAPH'22.
\end{itemize}
\ \\

@@

\bighead{LEAD CLINICAL RESEARCH -- ADVANCED CLINICAL IMAGING TECHNOLOGY}{Siemens Healthineers / Lausanne University Hospital (CHUV), Switzerland}{2011--2017}
\nopagebreak
\begin{itemize}
\item Led algorithmic development of computer-aided diagnosis software (released by Siemens Healthineers as part of the {\em AI-Rad Companion} solution) using brain morphometry for radiological reading. {\em Keywords:} magnetic resonance imaging (MRI), brain image segmentation, disease classification, neurodegeneration, Alzheimer's disease.
\item Main organizer of annual Siemens/CHUV {\em brain imaging} workshops in Lausanne, Switzerland, 2012--2016.
\end{itemize}

@@

\begin{tabular}{p{0.2\linewidth}p{0.8\linewidth}}
  \textbf{Languages:} & Native French, fluent English, basic Portuguese \& Italian.\\
  %\hline
  \ \\
  \textbf{IT skills:} &
  \tabitem Programming languages: Python, C, C++, Matlab, R.\\
  & \tabitem Long-term experience with scientific Python packages (numpy, scipy, pylab).\\
  & \tabitem CV/AI Python packages: skimage, PIL, opencv, sklearn, pytorch, tensorflow, keras.\\
  & \tabitem C/Python integration via Cython.\\
  & \tabitem Version control software: git, svn, perforce.\\
  & \tabitem Agile workflow (mainly Scrum) using JIRA \& Confluence.\\
  & \tabitem Reporting: \LaTeX, MS~Office, OpenOffice, LibreOffice, Google~Docs.\\
  %\hline
  \ \\
  \textbf{Publications:} &
  \tabitem ORCID: \url{http://orcid.org/0000-0002-4821-6893}.\\
  & \tabitem 116 scholarly publications.\\
  & \tabitem Google scholar statistics: 7603 citations, h-index: 35, i10-index: 74 (May 2023).\\
  & \tabitem 6 published international patents. 2 ongoing applications.\\
  %\hline
  \ \\
  \textbf{Academic work:} & 
  \tabitem Post-graduate teaching in computer vision and medical image analysis: about 100~hours (France: Ecole Centrale Paris, University of Nice, INSERM/CNRS continuing education; Switzerland: EPFL Lausanne, University of Lausanne).\\
  & \tabitem Long-time scientific journal reviewing experience (IEEE Trans.~Medical Imaging, Medical Image Analysis, NeuroImage, IEEE Trans.~PAMI, IEEE Trans.~Signal Processing, Frontiers in Neuroscience, \ldots).\\
  & \tabitem Supervision of two PhD students and four MSc students as main supervisor, three PhD students and two MSc students as co-supervisor.\\
  %\hline
  \ \\
  \textbf{References:} & Available on request.
\end{tabular}

### 2024-02-06 papers/variana/notes.tex
Work on CV

\def\x{\mathbf{x}}
\def\y{\mathbf{y}}
\def\e{\mathbf{e}}
\def\th{{\boldsymbol{\theta}}}
\def\n{{\boldsymbol{\nu}}}
\def\u{\mathbf{u}}
\def\m{\mathbf{m}}
\def\v{\mathbf{v}}
\def\f{\mathbf{f}}
\def\E{\mathbb{E}}
\def\g{\mathbf{g}}
\def\h{\mathbf{h}}
\def\A{\mathbf{A}}
\def\B{\mathbf{B}}

@@

\section{2023 thoughts}

Context: ill posedness (in the sense of Hadamard) of empirical risk minimization, including  maximum likelihood, MAP, etc. These methods are at the heart of statistical inference and machine learning, still they are based on ill-posed mathematical problems! Shouldn't we worry about that?

The empirical risk is just a proxy for a Bayesian risk that we are not able to compute unless we know the {\em true} distribution of the training data. We don't really mean to optimize the empirical risk.

Bayesian inference boils down to a problem of the form:
$$
r(\theta) = \int p(x) L(x, \theta) dx 
$$
where $p(x)$ is the distribution of some variable of interest and $\theta$ is the parameter to be learned. Because we don't know $p(x)$ in agnostic learning context, or we know it but it's intractable, we may resort to an empirical version of the risk:
$$
\hat{r}(\theta) = \frac{1}{n} L(x_i, \theta).
$$

@@

I have been bumping into this issue for years... only to realize now (January 2022) that such an upper bound does exist and is actually pretty simple to exhibit. We will need a straightforward intermediate result:

@@

Integrating this over~$\pi$, we get this long-waited inequality:
$$
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g).
$$

Note that this is only tight if $\max(f,g)$ is constant,

This tells us that, as intuition suggests, the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes. But this holds under the mild condition that both $f$ and $g$ remain bounded. This practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.

@@

The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi$. The above result implies straighforwardly:
$$
(f-g)\phi^2
\leq 2 \max(f\phi, g\phi) \ell(f\phi, g\phi)
= 2 \max(f\phi, g\phi) \phi\ell(f, g)
= 2 \max(f\phi^2, g\phi^2) \ell(f, g)
$$
where we use the fact that $\ell(f\phi,g\phi)=\phi\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
$$
\int \pi (f-g)^2\phi^2 \leq 2\max(\|f\phi^2\|_\infty, \|g\phi^2\|_\infty) D(\pi f\|\pi g).
$$

@@

>>>>

\def\x{\mathbf{x}}
\def\y{\mathbf{y}}
\def\e{\mathbf{e}}
\def\th{{\boldsymbol{\theta}}}
\def\n{{\boldsymbol{\nu}}}
\def\u{\mathbf{u}}
\def\m{\mathbf{m}}
\def\v{\mathbf{v}}
\def\f{\mathbf{f}}
\def\E{\mathbb{E}}
\def\g{\mathbf{g}}
\def\h{\mathbf{h}}
\def\A{\mathbf{A}}
\def\B{\mathbf{B}}
\def\prec{\mathrm{prec}}

@@

\section{History}

It is time to assemble the pieces of the puzzle.
\begin{itemize}
\item 2010: discovery of the VS core idea and its intriguing exactness property.
\item 2011--13: consolidation of the theory behind VS.
\item 2014--16: realization that VS is useful in practice combined with Expectation Propagation.
\item 2018--19: extension to large dimension via custom optimizer (avoiding big Hessian inversion).
\item 2022: proven bound on VS asymptotic variance.
\item 2023: final key insight: VS makes optimization possible in high dimension (at least locally) because it avoids saddle points unlike gradient-based methods.
\end{itemize}

\section{FAQ}

\paragraph{Is VS better than IS?} No. VS is asymptotically equivalent to IS applied to the residual Gaussian fitting error, {\em i.e.}, the difference between the target distribution and its KL optimal Gaussian fit. VS will thus be more accurate than IS whenever the residual importance weights vary less than the base importance weights in the sampling distribution support. This happens when the target is locally well approximated by a Gaussian. But this is not always the case. For instance, assume we want to estimate the normalizing constant of a distribution that we are able to sample from; in this case, the base importance weights are constant, and substracting the Gaussian fit doesn't help. VS doesn't even converge if the target has shorter tails than a Gaussian, while IS is exact no matter what. Generally speaking, VS doesn't appear as the method of choice when we have a good sampler.

%\paragraph{What does ``nearly Gaussian'' mean?} Roughly speaking, it means that the distribution can be decomposed as the product of a Gaussian distribution and a slowly varying function within the Gaussian distribution's support.

\paragraph{Is VS better than Laplace?} The Laplace method has systematic bias, unlike VS. This is fundamentally because it works at fixed, infinitesimal scale, while VS works at adjustable scale. One can think of simple situations where Laplace fails miserably, such as fitting a Laplace distribution (isn't that ironical?), but it may also work very well provided that the target is nearly Gaussian. The problem is that it is impractical in high dimension since it requires maximizing the distribution (which is more or less the problem we are trying to solve using VS). In EP context, we can perform a second-order Taylor expansion of the log-factor at the cavity center, yielding a damped Newton iteration. The resulting approximate EP is has a tendency to get stuck at saddle points, just like most gradient-based methods. VS-based EP is able to escape saddle points by exploring the target distribution in neighborhoods of increasing sizes.

\paragraph{Can I choose any sampling kernel in practice?} Unfortunately, no.

@@

\subsection{VS variance inequality}

I have been bumping into this issue for years... only to realize in January 2022 that such an upper bound does exist and is actually pretty simple to exhibit. We will need a straightforward intermediate result:

@@

Integrating this over~$\pi$, we get this long-waited inequality:
$$
\int \pi (f-g)^2 \leq 2\max(\|f\|_\infty, \|g\|_\infty) D(\pi f\|\pi g),
$$
where the infinite norms are taken on the support of~$\pi$, which is infinite if $\pi$ is a Gaussian kernel.

Note that this is only tight if $\max(f,g)$ is constant. The inequality tells us, as intuition suggests, that the mean squared error between factors smoothly goes to zero as the KL~divergence vanishes. But this holds under the condition that both~$f$ and~$g$ remain bounded on the support of~$\pi$. While this may be problematic if~$\pi$ has infinite support, we can just truncate~$\pi$ to guarantee the finiteness of the upper bound at the price of some possibly small bias. As noted in the 2013 paper, this boils down to using a localized version of the KL divergence as the target metric. It may also be justified in terms of a Bayesian prior, and may be a very reasonable thing to do. However, while removing the concern that the VS variance may be infinite, it doesn't prevent it in practice from being very large and yielding
hopelessly slow convergence.

The above inequality gives a variance upper bound on the zero-order moment (the normalizing constant). More generally, consider a positive real $\phi_i$. The above result implies straighforwardly:
$$
(f-g)\phi_i^2
\leq 2 \max(f\phi_i, g\phi_i) \ell(f\phi_i, g\phi_i)
= 2 \max(f\phi_i, g\phi_i) \phi_i\ell(f, g)
= 2 \max(f\phi_i^2, g\phi_i^2) \ell(f, g)
$$
where we use the fact that $\ell(f\phi_i,g\phi_i)=\phi_i\ell(f,g)$. A more general inequality follows by integration, which I would happily call Alexis' inequality:
$$
\int \pi (f-g)^2\phi_i^2 \leq 2\max(\|f\phi_i^2\|_\infty, \|g\phi_i^2\|_\infty) D(\pi f\|\pi g).
$$

\subsection{Sufficient conditions for VS convergence}

From this point, can we give practical sufficient conditions for VS convergence?  Convergence in distribution will occur under mild conditions if the integral $\int \pi (f-g_\star)^2\phi\phi^\top$ is finite. The above inequality relating the VS variance to the KL~divergence tells us that this holds whenever $\pi$ has finite support.

What if~$\pi$ is stricly Gaussian, hence doesn't have finite support?

Similarly to IS, the condition that $f$ is upper bounded is sufficient for $\pi f^2 \phi\phi^\top$ to be finite, and may be easy to fulfill in practice.

We are left with another sufficient condition (which IS doesn't have): that $\pi g_\star^2 \phi\phi^\top$ be finite too. Due to the log-quadratic nature of both $\pi$ and $g_\star$, this holds iff $\pi g_\star$ is Gaussian (i.e. its hegated log is a positive definite quadratic form).

In 1D, let us denote $s^2$ and $\sigma^2$ the respective variances of $\pi$ and $\pi g_\star$. Clearly, $\sigma^2$ is finite because it is equal to the variance of $\pi f$ by the moment matching constraint, which is finite if $f$ is upper bounded. A simple asymptotic analysis then shows that the condition that $\pi g_\star^2$ has finite integral boils down to:
$$
\prec(\pi) + 2 [\prec(\pi f) - \prec(\pi)] > 0,
\quad \Rightarrow \quad
\prec(\pi) < 2 \prec(\pi f),
$$
where $\prec$ denotes the precision, {\em i.e.}, the inverse variance. The condition boils down to:
$$
s^2 > \frac{\sigma^2}{2}.
$$

The same reasoning can be extended in multiple dimension. If we use factorized Gaussians for fitting, the above condition simply applies to each axis.

In short, the kernel size cannot be too small compared to the target distribution. However, we don't know $\sigma$, so how can we guarantee this in practice?

Let now consider a damping factor $0\leq \alpha \leq 1$, and raise our factor~$f$ to the power~$\alpha$.

is practically rules out the naive application of VS to unbounded importance weight functions, hence the need to perform VS within a constrained framework such as Expectation Propagation.

@@

@@

\section{Some 2023 thoughts}
